Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-10 paper.tex
Pass at intro. Muddled narrative, but improving.

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Three classes of ARG inference method have emerged as the focus of recent
work, and have all been successfully applied in different contexts. 
Parsimony
based methods~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022}
seek to minimise some aspect of the inferred ARG
(usually the number number of recombinations required), and work well for 
answering specific detailed questions about 
recombination~\citep[e.g.][]{ignatieva_ongoing_2022} when the expected
number of ancestral recombinations is small (tens to hundreds).
The second class of inference method is based on sampling ARGs statistically 
from a population genetics model~\citep{rasmussen_genome-wide_2014,
hubisz_mapping_2020,mahmoudi_bayesian_2022, deng_robust_2024}, and have 
been applied in several
contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. These methods are well-suited to applications 
in which population genetics models are a good fit [TODO: Say this differently, 
I just want to distinguish from e.g. SARS-CoV-2 where pop gen models are 
basically meaningless]
in cases in which the expected number of recombinations is in the 
thousands to tens-of-thousands range (e.g., human datasets in which
we have X samples and a segment of Y Mb of chromosome [TODO quantify
from SINGER paper). 
The third class of inference method and based on heuristics
and approximate models and focus on large scale inference,
where the expected number of recombinations is in the millions or
tens of millions [TODO check numbers here]~\citep{kelleher_inferring_2019,
speidel_method_2019,zhang_biobank-scale_2023}.
Other methods specialised to particular applications have also 
recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021}.
See~\cite{wong_general_2023} for a detailed review of ARG inference 
methods.

Typically, the more rigorous approaches can handle 10-100 samples,
while the approximate methods can scale to 10s to hundreds of
thousands of samples. With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference.

Perhaps the clearest route to statistical rigor is through scalable
model-based inference, a central element of population genetics.
The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
tajima_evolutionary_1983} has proved to be a useful model, and is the
basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
The incorporation
of recombination in the coalescent process~\citep{hudson_properties_1983}
has proved to be much more difficult.
% FIXME rephrase, next two sentences, lifted from ARG paper
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
nielsen_estimation_2000,
fearnhead_estimating_2001}.
% TODO reuse some of the ideas from the SMC paper here, it has
% some good explanations.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The introduction of the Sequentially Markovian Coalescent (SMC)
approximation to the coalescent with recombination
\citep{mcvean_approximating_2005} 
and its refinement in the SMC'~\citep{marjoram_fast_2006},
are the basis of many powerful inference 
% TODO add some more recent stuff
methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
In particular, the SMC is the basis for ARG sampling methods
like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
SINGER~\citep{deng_robust_2024}, which are the gold-standard 
in terms of inference accuracy.
ARGweaver has consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and, while SINGER has not been evaluated as thoroughly yet, it promises
to be even more accurate as well as substantially more scalable 
computationally~\citep{deng_robust_2024}.

% THE PROBLEM!!! OMG it's terrible!!!
Despite these recent advances in statistically rigorous ARG 
inference, however, there is still a gap of some three orders 
of magnitude between the sample-sizes that can be handled 
(hundreds) compared to the large-scale hueristic methods
like tsinfer and ARGneedle (hundreds of thousands).
[TODO working through a more positive narrative here, based on
the idea of combining the strengths of the approaches]
Although major breakthroughs in scale have been achieved by hueristic
ARG inference
methods~\citep{kelleher_inferring_2019,speidel_method_2019,zhang_biobank-scale_2023},
these methods are currently fundamentally incompatible with rigorous
inference under population genetic models such as the SMC. This is 
because these methods return approximate ARG
\emph{structures}~\citep{wong_general_2023} [WEAK, FIXME]. 
Specifically,
we cannot evaluate a likelihood of these ARGs in coalescent 
models under the classical 
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present
[Also: \cite{guo_recombination-aware_2022}]
[NOTE the following is super weak, and needs a lot of work.]
This inability to connect the results of large-scale inference methods
with a generative model
is a fundamental limitation, and limits both the utility of 
current point estimates as well as the possibilities for further
inference. Specifically, we are unable to quantify uncertainty 
around heuristic point estimates. We cannot explore ARG space
around the point estimates using (e.g. MCMC), as we have no way definition
of improvement. We cannot compare models of demography, or evaluate the 
goodness of fit for model estimates.

%% This is how we solve it
[TODO figure out the gARG, tree sequence etc mess]
Here, we bridge this gap between models and the products of approximate
inference by defining a likelihood function for any 
``genome ARG''~\citep{wong_general_2023} under the SMC. Using a novel
formulation of the SMC that focuses on the contributions of nodes and 
edges [FIXME] in a backwards-in-time way, we provide an efficient 
algorithm that can compute a likelihood for any gARG under this model.
We demonstrate the flexibility of this approach by XXX, and show
[SOMETHING].

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Calculating likelihoods} \label{par:liks}

@@

>>>>

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Although there are many different approaches to ARG
inference~\citep{wong_general_2023}, two broad classes of method
have been the focus of most recent interest.

% TODO Maybe swap the order of these two paragraphs around, so it's 
% more natrual to talk about the advantages of model-driven stuff next.
The first class of inference methods are statistically rigorous approaches
that sample ARGs under a population genetics model such as the
coalescent with recombination~\citep{hudson_properties_1983},
and its approximatation, the Sequentially Markovian
Coalescent~\citep{mcvean_approximating_2005,marjoram_fast_2006}.
ARGweaver~\citep{rasmussen_genome-wide_2014} is the most popular,
consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
citations,anywhere ARGweaver has been tested for accuracy]
and has succesfully been applied in several
contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. 
ARGweaver works [by explanation.]
The recently-introduced SINGER method takes a similar approach [doing something 
slightly different]~\citep{deng_robust_2024}.
While SINGER has not been evaluated as thoroughly as ARGweaver yet, it promises
to be even more accurate while being substantially more
scalable~\citep{deng_robust_2024}.

The second class of inference method that has been of recent interest 
is based on heuristics and approximate models.
Relate~\citep{speidel_method_2019},
tsinfer~\citep{kelleher_inferring_2019} and 
ARG-Needle~\citep{zhang_biobank-scale_2023} work on quite different 
principles, but share some common properties. Firstly, they are all 
hueristic and approximation driven, favouring computational
scalability over statistical rigor, and lacking a direct 
connection to population genetic theory. 
Secondly, they all produce a single
ARG as output, inferring a deterministic point estimate with no means 
of quantifying uncertainty. Thirdly (and most importantly for the 
purposes of this paper) the ARGs they produce lack substantial
detail about ancestral recombination events: the estimates are
rough sketches, hopefully capturing large-scale properties, but lacking
in the finer detail of outputs of ARGweaver or SINGER.

There are substantial advantages to the statistically rigorous sampling 
approaches of ARGweaver and SINGER. Firstly, they tend to be more 
accurate; ARGweaver has consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
citations,anywhere ARGweaver has been tested for accuracy]
and, while SINGER has not been evaluated as thoroughly yet, it promises
to be even more accurate~\citep{deng_robust_2024}. 
Secondly, the direct connection with
population genetic models greatly aids interpretability, and allows for
natural extensions like the joint estimation of ARGs and
demography~\citep{hubisz_mapping_2020}. Thirdly, sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates, which can be used in downstream applications [citation
of this?].

[Needs more work]
Despite these significant advantages over hueristic methods, however,
hueristic methods remain of interest because of the profound difference
in computational scalability between these approaches. Even with the 
major advances made in this direction by SINGER, ARG inference is 
only feasible for hundreds of samples. Tsinfer and ARG-Needle have 
been applied to datasets of hundreds of \emph{thousands} of samples.
With the explosive growth in dataset
sizes in recent years~\citep[e.g.][]{bycroft_genome_2018,halldorsson_sequences_2022,
all_genomic_2024} there is a pressing need for more statistical
rigor in large scale ARG inference. 
% This inability to connect the results of large-scale inference methods
% with a generative model
% is a fundamental limitation, and limits both the utility of 
% current point estimates as well as the possibilities for further
% inference. Specifically, we are unable to quantify uncertainty 
% around heuristic point estimates. We cannot explore ARG space
% around the point estimates using (e.g. MCMC), as we have no way definition
% of improvement. We cannot compare models of demography, or evaluate the 
% goodness of fit for model estimates.
A key problem facing the field is that the ARGs estimated by heuristic large-scale 
methods cannot be connected to population genetics theory because 
we cannot compute their likelihood under the SMC. The classical 
method~\citep{kuhner_maximum_2000} assumes complete knowledge about
recombination events, and there is no clear way to compute a likelihood for 
the approximate structures output by hueristic methods.

In this paper we provide a necessary first step towards this goal
of statistically rigorous large-scale ARG inference, by defining a function
to compute the likelihood of a general class of ARG under the SMC.
To do this, we rederive the SMC under a backwards-time formulation,
modelling the process in terms of lineages and [something] rather
than events. Using this novel formulation, we then show how to 
easily compute the likelihood of any ARG under the SMC and outline 
an efficient algorithm to do so. We illustrate the flexibility of this
approach via [some examples] and conclude by discussing avenues
for future research opened up.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Calculating likelihoods} \label{par:liks}
[Paragraph setting the scene on likelihood calculations]
\cite{kuhner_maximum_2000} formulation, as these formulations assume
that complete information about recombination events is present
[Also: \cite{guo_recombination-aware_2022}]
[NOTE the following is super weak, and needs a lot of work.]

@@

[Note: a point to make here is that the idea of combining hueristic large-scale 
inference with model based inference is analogous to the combination of 
coalescent simulation and forwards time simulation. We can also draw lessons
from developers of different methods cooperating and using the tskit library
to enable this.]
